---
title: "Association Between Sleep Duration and Cognitive Decline Among Older Adults"
author: "Mia LaBrier"
date: "`r Sys.Date()`"
output: html_document
---

# Load packages and open libraries
```{r}
pacman::p_load(odds.n.ends, blorr, lmtest, car,  broom, tidyverse, jtools, readr, haven, labelled, ggplot2, broom, table1, DiagrammeR) 
```

# Import BRFSS 2020 data
```{r}
setwd("/Users/mialabrier/Documents/advanced_data")
BRFSS2020 <- read_xpt("LLCP2020.XPT ")
```

# Clean data
## Get Information about imported data
```{r}
# Check variable types in dataframe
str(BRFSS2020)

# Get head of dataframe
head(BRFSS2020)
```

## Create new dataframe with selected variables for analyese
```{r}
# Select variables for analyses
BRFSS_2 <- BRFSS2020 %>%
  select(c("SLEPTIM1", "CIMEMLOS", "_AGEG5YR", "_SEX", "_IMPRACE"))

# Get head of new dataframe with selected variables
head(BRFSS_2)

# Rename selected variables
BRFSS_2 <- BRFSS_2 %>%
  rename("sleep_time" = "SLEPTIM1", "cognitive_decline" = "CIMEMLOS", "age" = "_AGEG5YR", "sex" = "_SEX", "race" = "_IMPRACE")

# Get head of new dataframe with renamed variables
head(BRFSS_2)
```

## Get information about new dataframe and variables selected for analyses
```{r}
# Check variable types for selected variables
str(BRFSS_2)

# Get an idea of values and NAs for selected variables
summary(BRFSS_2)
##NOTE: cognitive decline has 334,120 missing values because it was only asked to participants ≥45 and not asked across all states

# Remove attributes from dataframe
var_label(BRFSS_2) <- NULL
```

## Recode Variables for analyses
### Cognitive decline (outcome variable) as binary
```{r}
# Cognitive decline is indicated by a participants response to "During the past 12 months, have you experienced confusion or memory loss that is happening more often or is getting worse?" (Yes, No, Don't Know, Refused, Not asked/Missing)

# Create table for cognitive decline before recoding
table(BRFSS_2$cognitive_decline, useNA = "always")

# Recode cognitive decline as a binary variable, with 0 = No and 1 = Yes
BRFSS_2 <- BRFSS_2 %>%
  mutate(cognitive_decline_binary = case_when(cognitive_decline == 2 ~ 0,
                                              cognitive_decline == 1 ~ 1),
      cognitive_decline_binary = factor(cognitive_decline_binary, levels=c(0:1), labels=c("No cognitive decline", "Cognitive decline"))) 

# Create table to make sure the recode worked as expected
table(BRFSS_2$cognitive_decline, BRFSS_2$cognitive_decline_binary, useNA="always")
```

### Sleep time (exposure of interest) as categorical
```{r}
#Sleep time is indicated by a participants response to "On average, how many hours of sleep do you get in a 24-hour period?" (Numeric value, Don't know, Refused, Missing)

# Create table for sleep time before recoding
table(BRFSS_2$sleep_time)
##NOTE, 77 = "Don't Know", 99 = "Refused" - These responses will not be coded into categorical levels, rather, they will count as NAs. Therefore the highest categorical level will be set from 10 hours ≤ 24 hours

# Recode sleep time as a categorical variable
BRFSS_2 <- BRFSS_2 %>%
  mutate(sleep_time_cat = case_when(sleep_time == 7 ~ 0,
                                    sleep_time <= 4 ~ 1,
                                    sleep_time == 5 ~ 2,
                                    sleep_time == 6 ~ 3,
                                    sleep_time == 8 ~ 4,
                                    sleep_time >= 9 & sleep_time <= 24 ~ 5),
         sleep_time_cat = factor(sleep_time_cat, 0:5, c("7 (reference)", "≤4", "5", "6", "8", "≥9")))

# Create table to make sure the recode worked as expected
summary(BRFSS_2$sleep_time_cat)
```

### Race (covariate) as catagorical
```{r}
# Create table for race before recoding
summary(BRFSS_2$race)

# Recode race as a categorical variable, with 45-64 = 1 and 65-≥80 = 2
BRFSS_2 <- BRFSS_2 %>%
  mutate(race_cat = case_when(race == 1 ~ 1,
                                    race == 2 ~ 2,
                                    race == 3 ~ 4,
                                    race == 4 ~ 4,
                                    race == 5 ~ 3,
                                    race == 6 ~ 4),
         race_cat = factor(race_cat, 1:4, c("White", "Black", "Hispanic", "Other")))

# Create table to make sure the recode worked as expected
summary(BRFSS_2$race_cat)
```

### Recode age as binary
```{r}
# Create table for age before recoding
summary(BRFSS_2$age)

# Recode age as binary 
BRFSS_2 <- BRFSS_2 %>%
  mutate(age_cat = case_when(age == 6 ~ 1,
                             age == 7 ~ 1,
                             age == 8 ~ 1,
                             age == 9 ~ 1,
                             age == 10 ~ 2,
                             age == 11 ~ 2,
                             age == 12 ~ 2,
                             age == 13 ~ 2),
         age_cat = factor(age_cat, 1:2, c("45-64", "65-≥80")))

# checking to make sure recode worked
summary(BRFSS_2$age_cat)
table(BRFSS_2$age_cat)
```

### Recode sex as binary
```{r}
# Create table for sex before recoding
table(BRFSS_2$sex)

# Recode sex as binary, with Male = 0 and Female = 1
BRFSS_2 <- BRFSS_2 %>%
  mutate(sex_cat = case_when(sex == 1 ~ 0,
                             sex == 2 ~ 1),
         sex_cat = factor(sex_cat, 0:1, c("Male", "Female")))

# checking to make sure recode worked
summary(BRFSS_2$sex_cat)
```

## Drop NAs (complete case analysis)
```{r}
#Drop NAs, participants with any missing values in BRFSS_2 will be excluded from analyses)
BRFSS_ex <- BRFSS_2 %>%
  drop_na()

100 - nrow(BRFSS_ex)/nrow(BRFSS_2) * 100
##NOTE: This is high, however, most of these missing values come from the cognitive decline variable, which was only collected for adults of 45 in specific states
```

# Assumptions and analyses
##Check number of cases in each exposure/outcome category
```{r}
#Create a table to determine the number of cases in each exposure outcome
table(BRFSS_ex$cognitive_decline_binary, BRFSS_ex$sleep_time_cat, BRFSS_ex$race_cat, BRFSS_ex$age_cat, BRFSS_ex$sex_cat)
##NOTE: The assumption for a large enough sample size is not met (there are not at least 10 cases in each exposure/outcome category). However, condensing sleep and/or race categories further would not prohibit valuable comparison across these categories

table1(~ factor(sleep_time_cat) + factor(age_cat) + factor(sex_cat) + factor(race_cat) | cognitive_decline_binary, data=BRFSS_ex)
```
## Reformat Table 1 and create Figure 1
```{r}
#Table 1
label(BRFSS_ex$age_cat)<-"Age group (years)"
label(BRFSS_ex$sex_cat)<-"Sex"
label(BRFSS_ex$race_cat)<-"Race"
label(BRFSS_ex$sleep_time_cat)<-"Sleep duration (hours"
label(BRFSS_ex$cognitive_decline_binary)<-"Cognitive decline"

table1(~ sleep_time_cat + age_cat + sex_cat + race_cat | cognitive_decline_binary, overall="Total", rowlabelhead="Variable", BRFSS_ex)

#Figure 1
grViz("digraph flowchart {

      node [fontname = Helvetica, shape = rectangle, fontsize=10] 
      tab1 [label = '@@1'] # starting
      tab2 [label = '@@2'] # exclusion 1
      tab3 [label = '@@3'] # exclusion 2
      
      tab1 -> tab2 -> tab3
}

      [1]: 'Total BRFSS 2020 surveys n=401,958'
      [2]: 'Excluding 334,120 individuals who did not participate in Cognitive Decline module n=67,838'
      [3]: 'Excluding 2,411 individuals with missing data on variables used in the analysis n=65,427'
      ")
```

## Model 1: Univariate (unadjusted) logistic model with cognitive decline as the outcome and sleep time as the exposure
```{r}
# Univariate model, with cognitive decline as the  outcome and sleep time as the exposure
unadjusted_model <- glm(cognitive_decline_binary ~ sleep_time_cat, data=BRFSS_ex, family="binomial") 
summary(unadjusted_model)

# Exponated coefficients
exp(unadjusted_model$coefficients)
exp(confint(unadjusted_model))

# Odds Ratios and additional information
odds.n.ends(unadjusted_model) 
##NOTE: This model has 0 sensitivity (does not correctly classify any true positives) and 1 specificity (correctly categorizes all true negatives). All ORs significant.
```

## Model 2: Model with cognitive decline as the outcome and sleep time as the exposure, adjusted for age, race, and sex
```{r}

# Model with cognitive decline as the  outcome and sleep time as the exposure, adjusted for age, race, and sex
adjusted_model <- glm(cognitive_decline_binary ~ sleep_time_cat + age_cat + race_cat + sex_cat, data=BRFSS_ex, family="binomial") 
summary(adjusted_model)

# Exponated coefficients
exp(adjusted_model$coefficients)
exp(confint(adjusted_model))

# Odds Ratios and additional information
odds.n.ends(adjusted_model) 

# Compare unadjusted model to adjusted model
compareCoefs(unadjusted_model, adjusted_model)
lrtest(unadjusted_model, adjusted_model)
## NOTE: This model has 0 sensitivity (does not correctly classify any true positives) and 1 specificity (correctly categorizes all true negatives). All ORs re: sleep time and cognitive decline are still significant.

# Hosmer lemeshow goodness of fit test (a significant p value indicates a bad fit)
blr_test_hosmer_lemeshow(adjusted_model)
```

## Check for effect modification
### By age
```{r}
# Testing interaction term for sleep time*age, while controlling for other covariates)
age_interm_model <-glm (cognitive_decline_binary ~ sleep_time_cat + age_cat + race_cat + sex_cat + sleep_time_cat*age_cat, data=BRFSS_ex, family="binomial") 
summary(age_interm_model)

# Exponated coefficients
exp(age_interm_model$coefficients)
exp(confint(age_interm_model))

# Odds Ratios and additional information
odds.n.ends(age_interm_model)

# Test hypothesis (age modifies the association between sleep time and cognitive decline) with LRT
lrtest(adjusted_model, age_interm_model)
## NOTE: the interaction term is significant, stratify analyses by age groups
```

### By race
```{r}
# Testing interaction term for sleep time*race while controlling for other covariates)
race_interm_model <-glm (cognitive_decline_binary ~ sleep_time_cat + age_cat + race_cat + sex_cat + sleep_time_cat*race_cat, data=BRFSS_ex, family="binomial") 
summary(race_interm_model)

# Exponated coefficients
exp(race_interm_model$coefficients)
exp(confint(race_interm_model))

# Odds Ratios and additional information
odds.n.ends(race_interm_model)

# Test hypothesis (age modifies the association between sleep time and cognitive decline) with LRT
lrtest(adjusted_model, race_interm_model)
## NOTE: the interaction term is not significant, adjust for race (do not stratify)
```

### By sex
```{r}
# Testing interaction term for sleep time*sex while controlling for other covariates)
sex_interm_model <- glm(cognitive_decline_binary ~ sleep_time_cat + age_cat + race_cat + sex_cat + sleep_time_cat*sex_cat, data=BRFSS_ex, family="binomial") 
summary(sex_interm_model)

# Exponated coefficients
exp(sex_interm_model$coefficients)
exp(confint(sex_interm_model))

# Odds Ratios and additional information
odds.n.ends(sex_interm_model)

#test hypothesis (age modifies the association between sleep time and cognitive decline) with LRT
lrtest(adjusted_model, sex_interm_model)
## NOTE: the interaction term is not significant, adjust for sex (do not stratify)
```

## Model 3: Model with cognitive decline as the outcome and sleep time as the exposure, adjusted for race and sex
```{r}
# This was created just in case it would be helpful to compare an adjusted model (after determining age is NOT a confounder) to an unadjusted model, when NOT stratifying by age

# Model with cognitive decline as the outcome and sleep time as the exposure, adjusted for race and sex (not age, as age is a modifier not a confounder)
adj_sex_race_model <- glm(cognitive_decline_binary ~ sleep_time_cat + race_cat + sex_cat, data=BRFSS_ex, family="binomial") 
summary(adj_sex_race_model)

# Exponated coefficients
exp(adj_sex_race_model$coefficients)
exp(confint(adj_sex_race_model))

# Odds Ratios and additional information
odds.n.ends(adj_sex_race_model) 

# Compare unadjusted model to newly adjusted model
compareCoefs(unadjusted_model, adj_sex_race_model)
lrtest(unadjusted_model, adj_sex_race_model)
```
## Create stratified datasets (BRFSS_ex stratified byt age group)
```{r}
BRFSS_ex_45_64 <- subset(BRFSS_ex, age_cat == "45-64")
  summary(BRFSS_ex_45_64)
BRFSS_ex_65_80 <- subset(BRFSS_ex, age_cat == "65-≥80")
  summary(BRFSS_ex_65_80)
```


## Model 4: Univariate (unadjusted) model for 45-64 age group
```{r}
# Made for comparison purposes (i.e. to compare to adjusted model for 45-64 age group)

# Univariate model, 
unadjusted_model_45_64 <-glm (cognitive_decline_binary ~ sleep_time_cat, BRFSS_ex_45_64, family="binomial") 
summary(unadjusted_model_45_64)

# Exponated coefficients
exp(unadjusted_model_45_64$coefficients)
exp(confint(unadjusted_model_45_64))

# Odds Ratios and additional information
odds.n.ends(unadjusted_model_45_64) 

# Hosmer lemeshow goodness of fit test (a significant p value indicates a bad fit)
blr_test_hosmer_lemeshow(unadjusted_model_45_64)
```

## Model 5: Univariate (unadjusted) model for 65-≥80 age group
```{r}
# Made for comparison purposes (i.e. to compare to adjusted model for 65-≥80 age group)

# Univariate model, 
unadjusted_model_65_80 <-glm (cognitive_decline_binary ~ sleep_time_cat, BRFSS_ex_65_80, family="binomial") 
summary(unadjusted_model_65_80)

# Exponated coefficients
exp(unadjusted_model_65_80$coefficients)
exp(confint(unadjusted_model_65_80))

# Odds Ratios and additional information
odds.n.ends(unadjusted_model_65_80)

# Hosmer lemeshow goodness of fit test (a significant p value indicates a bad fit)
blr_test_hosmer_lemeshow(unadjusted_model_65_80)
```

## Model 6: Adjusted model for 45-64 age group
```{r}
# Model with cognitive decline as the outcome and sleep time as the exposure, adjusted for race and sex (45-64 age group)
model_45_64 <- glm(cognitive_decline_binary ~ sleep_time_cat + race_cat + sex_cat, BRFSS_ex_45_64, family = "binomial") 
summary(model_45_64)

# Exponated coefficients
exp(model_45_64$coefficients)
exp(confint(model_45_64))

# Odds Ratios and additional information
odds.n.ends(model_45_64)

# Compare unadjusted stratified model to adjusted stratified model
compareCoefs(unadjusted_model_45_64, model_45_64)
lrtest(unadjusted_model_45_64, model_45_64)

# Hosmer lemeshow goodness of fit test (a significant p value indicates a bad fit)
blr_test_hosmer_lemeshow(model_45_64)
```

## Model 7: Adjusted model for 65-≥80 age group
```{r}
# Model with cognitive decline as the outcome and sleep time as the exposure, adjusted for race and sex (65-≥80 age group)
model_65_80 <- glm(cognitive_decline_binary ~ sleep_time_cat + race_cat + sex_cat, BRFSS_ex_65_80, family = "binomial") 
summary(model_65_80)

# Exponated coefficients
exp(model_65_80$coefficients)
exp(confint(model_65_80))

# Odds Ratios and additional information
odds.n.ends(model_65_80) 

# Compare unadjusted stratified model to adjusted stratified model
compareCoefs(unadjusted_model_65_80, model_65_80)
lrtest(unadjusted_model_65_80, model_65_80)

# Hosmer lemeshow goodness of fit test (a significant p value indicates a bad fit)
blr_test_hosmer_lemeshow(model_65_80)

# Hosmer lemeshow goodness of fit test (a significant p value indicates a bad fit)
blr_test_hosmer_lemeshow(model_65_80)
```

## Tests for multicolliniarity for stratified adjusted models
```{r}
# Test for multicolliniarity for adjusted model for 45-64 age group
vif(model_45_64)

# Test for multicolliniarity for adjusted model for 65-≥80 age group
vif(model_65_80)
```

## Influential data points
### Test for influential data points for stratified adjusted models
```{r}
# Influential data points for adjusted model for 45-64 age group

## Plot Cook's Distance
plot(model_45_64, which = 4, id.n = 3, col="red")

## Cook's D calculations for each observation, plus other stats
model_45_64.data <- augment(model_45_64) %>%  
  mutate(index = 1:n())

head(model_45_64.data)

# Influential data points for adjusted model for 65-≥80 age group

## Plot Cook's Distance
plot(model_65_80, which = 4, id.n = 3, col="red")

## Cook's D calculations for each observation, plus other stats
model_65_80.data <- augment(model_65_80) %>%  
  mutate(index = 1:n())

head(model_65_80.data)
```

### Identify and remove influential data points
```{r}
# Identify and remove influential data points for adjusted model for 45-64 age group
## Define cutoff variable
cuttoff <- mean(model_45_64.data $.cooksd)*3

## Exclude observations
BRFSS_out_45_64 <- model_45_64.data  %>% # Exclude observations
  filter(.cooksd<cuttoff)

## Check percent kept
nrow(BRFSS_out_45_64)/nrow(BRFSS_ex_45_64)

# Identify and remove influential data points for adjusted model for 65-≥80 age group
## Define cutoff variable
cuttoff <- mean(model_65_80.data$.cooksd)*3

## Exclude observations
BRFSS_out_65_80 <- model_65_80.data %>% # Exclude observations
  filter(.cooksd<cuttoff)

## Check percent kept
nrow(BRFSS_out_65_80)/nrow(BRFSS_ex_65_80)

```

### Stratified adjusted model with influential data points excluded
```{r}
# Stratified adjusted model for adjusted model for 45-64 age group, influential data points removed
model_45_64_drop <- glm(cognitive_decline_binary ~ sleep_time_cat + race_cat + sex_cat, BRFSS_out_45_64, family = "binomial") 
summary(model_45_64_drop)

# Stratified adjusted model for adjusted model for 64-≥80 age group, influential data points removed
model_65_80_drop <- glm(cognitive_decline_binary ~ sleep_time_cat + race_cat + sex_cat, BRFSS_out_65_80, family = "binomial") 
summary(model_65_80_drop)
```

### Compare coefficients between models with and without influential observations
```{r}
#45-64
compareCoefs(model_65_80, model_65_80_drop) 

# 65-80
compareCoefs(model_45_64, model_45_64_drop)

## NOTE: Dropping influential data points results in complete separation and does some very weird things to both models! #Basically, I think it dropped everyone with cognitive decline, therefore, the model was able to predict *perfectly* (for the remaining observations)
```

### Check variables by cognitive_decline_binary
```{r}
#45-64
table(BRFSS_out_45_64$sleep_time_cat, BRFSS_out_45_64$cognitive_decline_binary)

# 65-80
table(BRFSS_out_65_80$sleep_time_cat, BRFSS_out_65_80$cognitive_decline_binary)

## NOTE: Removing all influential data points did, in fact, remove everyone who reported cognitive decline! Therefore, influential data points should not be removed.
```
```



